% \noindent Bias and fairness in machine learning are topics of considerable recent research interest~\cite{pedreshi2008discrimination,fairness,bozdag_bias_2013}.
% A standard approach in this area is to identify a variable or variables representing membership in a protected class, for example, race in an employment context, and to develop algorithms that remove bias relative to this variable. See, for example, ~\cite{zemel2013learning,kamishima2012fairness,kamiran2010discrimination,zhang2017anti}.

% To extend this concept to recommender systems, we must recognize the key role of personalization. Inherent in the idea of recommendation is that the best items for one user may be different than those for another. It is also important to note that recommender systems exist to facilitate transactions. Thus, many recommendation applications involve multiple stakeholders and therefore may give rise to fairness issues for more than one group of participants~\cite{abdollahpouri_recommender_2017}.


% intro of ofair
% Recommender systems are designed to assist users to find items of interest. Such systems model users' historical behaviors and generate personalized recommendations tailored to users' interests or needs. Recent research has identified a key limitation in a user-focused approach to recommender systems development, namely that it ignores multistakeholder aspects of the systems in which recommendation is embedded\cite{abdollahpourimulti2020}. In particular, the problem of \textit{provider fairness} has been underappreciated in recommender systems research, as it concerns the impact of recommendation delivery on the providers of items being recommended and the questions of fair treatment that may arise\cite{burke2017multisided}.

% Recent research has sought to alleviate this concern using a variety of approaches. See, for example, \cite{yao2017beyond,burke2018balanced,ekstrand2018exploring,liu2019personalized,kamishima2016model,beutel2019fairness}. What these approaches share is that they focus on a single dimension over which fairness is sought: a single protected group among the providers, and except for \cite{liu2019personalized}, they do not take user preferences in item features into account. % across item dimensions into account.

% The problem of promoting provider fairness while maintaining recommendation accuracy can be generally characterized as a multi-objective optimization problem. If optimal fairness and optimal recommendation accuracy could be achieved simultaneously, there would be no need for research in this area. However, optimizing recommendation accuracy often comes at the expense of provider fairness, due to various biases present in recommender systems, including popularity bias \cite{celma2008hits,lee2014fairness}, and user-base composition \cite{lin2019crank, yao2017beyond}. Research in provider fairness is therefore generally concerned with improving the tradeoff between fairness and accuracy, or in other words, increasing the amount of fairness that can be gained for a given degree of accuracy loss.

% A standard approach in this area is to identify a variable or variables representing membership in a protected class, for example, race in an employment context, and to develop algorithms that remove bias relative to this variable. See, for example, ~\cite{zemel2013learning,kamishima2012fairness,kamiran2010discrimination,zhang2017anti}.


% ----------------------------------------------------------------------------------------
% ----------------------------------------------------------------------------------------

% outline
% why fairness?
% what is fairness in machine learning? The problem that existed, Cambridge Analytica and those stories, Facebook, etc.
% what fairness means in machine learning? how fairness is detected in machine learning? What are the solutions presented?
% How this problem extends to other areas of machine learning such as recommendation system?
% what is recommender system? why these solutions aren't all applicable to recsys?
% what are the potential problems in them?
% what is fairness in recsys?
% how fairness is detected in recsys?
% what are the solutions presented?

how can we group these measures? exposure-based and accuracy/error based that are trying to achieve some sort of equity or equality between subgroups.
What are the research questions from all the papers that we were curious about?
in general, all the following solutions are trying to find a balance between accuracy and fairness. 

What are the solutions offered?
- through regularization - balanced neighborhood
- through re-ranking opportunistically through opportunistic fairness.
    - pfar/ofar (one dimensional)
    - ofair (multi dimensional and defining opportunistic fairness (your loss is my gain))
    - dynamic fairness (fairness over time and the social choice and reranking)
    - future work

What tools did we design? and how did we contribute?

 
% ----------------------------------------------------------------------------------------
% ----------------------------------------------------------------------------------------
% motivation
After a period of substantially fast development in different aspects of digital systems, and with more and more decisions being delegated to algorithms, the society begins to realize these systems that were intended to assist people in different tasks have ethical issues and can cause harm to individuals and the society\todo{cite the cambridge analytica}. Discrimination caused by algorithms that are trained on bias data or by lack of a good design, propagation of bias\todo{cite}, marginalization of minority groups in the society, inflation in the polarization of the society that can be caused by tight filter bubbles, and etc. are among the harms that algorithms can potentially cause.\todo{cite} These problems have gained the attention of a multidisciplinary community from computer scientists, social scientists and legal scholars. Thus as a response, recent research has shifted from design of algorithms that pursue purely optimal outcomes with respect to a fixed objective function into ones that also consider social impacts such as fairness.
% fairness in ML
Fairness, bias and discrimination are topics of considerable research interest in the recent years~\cite{pedreshi2008discrimination,fairness,bozdag_bias_2013}.

% what fairness means in machine learning? how fairness is detected in machine learning? What are the solutions presented?
Much of the work in algorithmic fairness has been focused on classification methods with a myriad of definitions that has been proposed\todo{cite}. The key definition are explained in \cite{mitchell2021algorithmic}.

One of the main divisions in fairness definitions comes from the way we assess and evaluate fairness and whether this evaluation is individual based or group based.

%group fairness
In group fairness based definitions, a model’s treatment of two groups with respect to a sensitive attribute (e.g. gender, race, ethnicity, etc.) is compared. In this method, the protected group(s) is designated with respect to a previously defined sensitive attribute and should be protected against discrimination. The sensitive attribute definition is usually rooted in anti-discrimination laws\cite{barocas2016big}.This notion of fairness tries to ensure that algorithms don't impact the members of the protected group more adversely and disproportionately. Group-based methods of fairness has helped build the most prevalent structures to achieve and assess fairness ~\cite{zemel2013learning,kamishima2012fairness,kamiran2010discrimination,zhang2017anti}.

    % Statistical Parity
    % This notion has become the most prevalent structure, but even there researchers have shown tension between different definitions \todo{cite}.
    The notion of Statistical or Demographic Parity requires that two groups with a different sensitive group have the same chance of getting a positive result. This notion has been discussed under different names of avoiding disparate impact \cite{Feldman2015}, independence \cite{barocas2018fairness} and anti-classification \cite{corbett2018measure}. This measure is used to ensure a ``fair'' representation of different groups in different tasks such as ranking\cite{singh2018fairness,zehlike2017fa,yang2017measuring}, and recommendation\cite{mehtora2018towards,ekstrand2018exploring}.
    
    % Performance Parity
    Performance Parity is another category of group fairness that requires equal error rates for different groups. Equality of opportunity or equality of true positive rates \cite{hardt2016equality} that requires the positive classification ratees to be independent of the protected attribute given the true label, equality of both true positive and false positive rates which is known as equalized odds \cite{hardt2016equality}, equality of misclassification rates (e.g. equality of false negative rates aka lack of disparate mistreatment\cite{zafar2017fairness}) and equality of positive predictive values aka calibration, belong to this category. 
    Performance Parity has also been studied as error parity in recommendation algorithms\cite{ekstrand2018all,yao_huang_fatml-2017}.


%individual fairness
Dwork et al. \cite{Dwork2012individual} observed that the demographic parity requirements can be met when qualified candidates from one groups and random candidates from the other groups were chosen. Thus, satisfying certain group based fairness notions might deteriorate fairness for individuals in a group. Therefore fairness might be a requirement on an individual level.
Dwork et al. \cite{Dwork2012individual} introduces the concept of individual fairness \cite{Dwork2012individual} which posits that similar individuals with respect to the task at hand should be treated similarly or in other words should have similar probabilities of positive classification outcomes. One of the limitations of this method is the choice of similarity metric to compare individuals and whether this metric is unbiased. This method also doesn't place any requirement on the treatment of dissimilar individuals.
Another individual fairness definition was proposed by \cite{pmlr-v70-kearns17a} for the problem of candidate set selection from diverse incomparable source sets. Choosing candidates for a research position from a diverse research communities with uncomparable research metrics (e.g. citation rates are different in different research communities) is an example of this problem. Meritocratic fairness requires that less qualified candidates are probabilistically almost never chosen over the more qualified candidates.

    
% harm!
Independent to group and individual fairness, Crawford \cite{crawford2017trouble} defines two new fairness definitions which are connected to the harm caused by unfairness: (a) distributional harm that is cause by an inequitable distribution of a resource or opportunities, and (b) representational harm, where a system doesn't have an accurate representation of the society or where it systematically misrepresents certain groups.


The motivation of fairness constructs also categorizes fairness definitions into two groups: anti-classification and anti-subordination. U.S. anti-discrimination law is rooted in anti-classification ideas which requires that the influence of a protected group should not play a role in the decision making process. This notion can also be called as disparate treatment. The main idea o anti-subordination is to actively work to reverse the effects of historical discriminatory patterns in the decision making processes \cite{barocas2016big}.

Although these motivations are often very clear, their proper application is still vague \cite{xiang2019legal}. It is also worth mentioning that it has been shown that satisfying different fairness notions at the same time is mathematically impossible and infeasible \cite{Kleinberg:InherentTrade,chouldechova2017fair}. Due to the competing and sometimes conflicting goals of different fairness definitions, different needs of various stakeholders involved, etc. we cannot have a system that is universally "fair". Thus, it is essential to pick a fairness notion that serves best in the specific context of the target application. 


% what are recommender systems?
Recommender systems are one of the most pervasive applications of machine learning in industry. They play a pivotal role in connecting users to relevant items or content throughout the web while not only users rely heavily on them but also content producers, sellers or information providers.

% How this problem extends to other areas of machine learning such as recommendation system?
Traditionally the focus of recommendation algorithms have been on accuracy and it was known that it's tied strongly with user satisfaction. Later on, the focus of these systems changed to beyond accuracy methods such as diversity, coverage, novelty, serendipity. This change of focus was supported by the literature that showed these properties in recommendation lists of users increase their overall satisfaction \todo{cite}. In recent years, aligned with the change of focus on these beyond-accuracy and socially sensitive properties in machine learning, the social aspects of these algorithms has come to the fore.

% to the injustice they have caused for the society \todo{cite}.
Therefore achieving fairness in recommendation algorithms has gotten more attention. However, the goal of fairness isn't completely new in the recommender system literature. Alleviating the problem of popularity bias in recommendations \cite{popbias2018} and ensuring equality or equity in long-tail recommendations \cite{ferraro2019} can be thought of achieving fairness for content providers in the systems. Group recommendation also tries to recommend items to users while considering and treating all the members of the group fairly \cite{kaya2020}. However, the goal of fairness in recommendation goes beyond one stakeholder and is not bounded to the previously mentioned problems. Rather it focuses on the aspects that are socially sensitive such as discrimination against sensitive-groups, under-representation of sensitive groups and preventing biases from creeping into these systems.

% why recsys is different than machine learning
Recommender systems have their unique challenges for investigating the fairness concepts and the methods that have been developed in other machine learning literature is not fully applicable in recommender systems. The role of personalization and  multistakeholder nature of recommender systems add major additional complications to the problem of fairness in recommendations.
% ranking the outputs and the context add additional complications to the problem of fairness in recommendations.
    
    % personalization
    Penalisation interferes with the goal of fairness. This is mainly because fairness for all means that all the users in a system should receive the same recommendations  while the goal of personalization is to find the best item for each user which could be different for different users.
    
    % multistakeholderness
    recommender systems exist to facilitate transactions between consumers, content providers. Thus, many recommendation applications involve multiple stakeholders and therefore may give rise to fairness issues for more than one group of participants~\cite{burke_multisided_2017}. Consumer fairness, is concerned with fair and equitable treatment of all the users in the system regardless of their membership to any protected group. For example ensuring that all the subgroups of users are receiving quality recommendations not only certain groups. Provider fairness is concerned with a fair treatment of content providers or content creators. For example, by making sure they are represented in the recommendations fairly and have equal opportunities to benefit from the system. Subject fairness is concerned with fair treatment of the content, people or entities in a system. For example, ensuring that recommendations do not systematically under-represented specific segments of the society or certain content.
    Therefore, recommendation fairness is not one problem. Multiple stakeholders might be involved and although they might all seek fairness, their goal of fairness might be different (due to different definitions of fairness), competing or conflicting.
    
    % ranked outputs?
    % context of recommender systems?
    % what is fairness in recsys? we have already explained this before..
    
    % how fairness is detected in recsys? and this ...

As mentioned previously, we cannot achieve a universally fair system, therefore for each problem, it's essential that we consider the target stakeholders, the definition of harm or unfairness and the specific metrics for measuring harm or integrating in the system to avoid harm. These elements are important in order to define, integrate and assess fairness in recommendation algorithms.

% Besides the previously mentioned challenges, 

Lack of appropriate data to study fairness goals is another challenge with which we have to deal. We might need sensitive attributes (e.g. gender, race, etc.) for the stakeholder entities but sharing and using this data might have privacy, legal or ethical concerns. Some of solutions that were used for this problem were integrating different datasets, using inference methods to impute demographic information, generating synthetic datasets, or training algorithms without demographics. But, all of the previous solutions bring their specific limitations to the method. 

% fairness solutions and methods presented in recommender systems.
Fairness notions can be defined, assessed and integrated to the algorithms for different stakeholders. For each stakeholder here we investigate the prior work and categorize it based on previous fairness notions such as group fairness, individual fairness, etc.

        % consumer fairness
        % individual fairness
        % group fairness
        % fairness beyond accuracy
        % more complex scenarios

    % consumer side fairness
    Consumer fairness or (C-fairness) is concerned with the fair treatment of consumers (of recommendations) and the impact that recommendations have specifically on marginalized groups. Such objectives are sometimes required by law.
    %individual fairness
    In collaborative filtering, the recommendations are build based on the similarity of users. And the goal is to use these similarities to recommend items to users in a personalized way. In other words, the recommendations of similar users are similar, thus they are treated similarly. This property might look alike the definition of individual fairness, although the metric based on which the similarity is calculated is different. As an example
    Collaborative filtering uses the rating behavior of users whereas individual fairness looks at the user demographic information to calculate similarity which is hard to get by.
    Another issue is that while similar users will be treated similarly, but since the data is not rich for minority groups, statistically, all of the users in that group might be treated unfairly when compared to the whole.
    As an example, if women in a job recommendation platform tend to click on lower paying jobs, and since their rating behavior is similar, all of them get equally low paying jobs. Therefore, individual fairness and personalization might have similar goals, but the similarity metric and the information they use is different.
    % group fairness
    to integrate group fairness in recommendation algorithms, we can imagine the utility that the consumers receive from recommendations and whether it is distributed equally or whether all the individuals are benefiting equally from it. Similar to the group fairness definition in machine learning, here also we have both categories of (a) statistical or demographic parity and (b) performance parity when the first category focuses on sub-group representation while the latter focuses on subgroup loss (or gain). Since the performance parity is the statistical parity in performance metrics, it has been called statistical parity in previous research\todo{cite}.
    To achieve demographic parity, \cite{ekstrand2018exploring} ensures that group proportions in the recommendation sets should be similar to group proportions in input ratings. 
    Performance parity in recommendation is calculated based on the effectiveness of recommendations and whether different subgroups are experiencing the same accuracy or error. Although not all of the definitions of error in classification can be applied to recommendation settings. Similar to \cite{kamishima2016model}, Yao and Huang \cite{yao_huang_fatml-2017} have designed different error-based fairness metrics for collaborative filtering such as value unfairness, over-representation, under-representation, etc. They have compared the discrepancy between the actual and predicted ratings for protected and unprotected groups or inconsistencies between the predicted ratings for these groups. 
    \cite{ekstrand2018all} has performed an off-line top-N evaluation of several collaborative filtering algorithms and has compared the results of different user demographics based on their NDCG.
    \cite{steck2018calibrated} introduces the concept of miscalibration in recommendations which has been used detected as consumer unfairness. Miscalibration happens when the item preferences of the users in their profile isn't covered in the recommendations they receive. \cite{Kun2020calib} has discusses that usually smaller or niche subgroups receive more miscalibrated recommendations compared to bigger or more popular subgroups.
    % \cite{burke2018balanced} has also compared the statistical parity in precision for different demographic groups in consumers. 
    
        
    % provider fairness
        % provider utility
        % individual fairness
        % group fairness
        
        
    Provider-side fairness or (P-fairness) is concerned with treating the suppliers of information or items that are being recommended fairly. We can think of recommendation opportunities as a resource and the fair distribution of those opportunities among providers as provider fairness. 
    Much of the research on diversity and decreasing popularity bias, contributes to provider fairness. Although, provider fairness derives from the goal of social justice and promotes the content from the underprivileged groups to provide more opportunities for them to be discovered. 
    To achieve P-fairness, \cite{mehtora2018towards} tried to ensure that different sub-groups are similarly represented in the recommendations.
    
    The utility defined in this context is mostly defined as exposure. A recommendation list is a short list that provides limited opportunities to expose items to consumers. According to the user attention pattern, the items that are on top of the list, receive more attention and this attention decreases as we go down in the list. Therefore, not all the positions in a list have equal utilities and only one item in the whole list can have the most valuable position and receive the most benefit \cite{diaz2020}. Therefore provider utility is calculated over all the recommendation lists delivered to all the users, while to calculate each consumers utility we only need to look at each consumer's recommendation and nothing more. Most of the metrics introduced for provider utility are based on NDCG \cite{biega2018equity}.\todo{cite}
    
    %individual fairness
    individual fairness for providers mean that similar items should receive similar utility from the recommender system. As an example, items that bring the same utility to a user (the user likes them both), are considered similar and should receive the same exposure in a recommendation list. \cite{biega2018equity} aggregates each item's attention and relevance over multiple rankings and assumes the providers are being treated fairly if the attention they have received from users are proportional to their relevance. The similarity of providers can be calculated in many other different ways and is an open research area.
    
    %group fairness
    group fairness for providers is concerned with a fair treatment of different provider sub-groups. Although this goal should be aligned with the goal of personalization which considers users preferences so it doesn't recommend provider groups to users that don't like them. \cite{kamishima2018recommendation} proposes that the recommendations outcomes should be statistically independent of a sub-group's protected attribute in order to have group fairness. In this case, the probability that an item shows up in a recommendation list is independent of its sensitive attribute.
    Another method is to ensure that there is a fair representation of providers in the recommendation lists. Unfairness in this case is when there is a big divergence between the distribution of the provider groups in the lists and the target distribution. \cite{yang2017measuring,das2019conceptual}. This divergence can be calculated using KL-divergence, difference in probabilities, odds ratio, etc. \cite{biega2018equity} calculates the provider groups fairness by aggregating expected exposure over multiple rankings. Fairness in this context happens when each provider group receives an appropriate level of exposure. \cite{beutel2019fairness} incorporates a fair construct (disparate mistreatment) in BPR\cite{rendlebpr2009} loss function. In this construct, a fair ranking happens when ranking a relevant item over an irrelevant item is independent of its group membership. Their pairwise fairness objective is defined in two way: once between groups and once within groups.
    
    % subject fairness
    % other stakeholders
    besides the consumers and providers, there might be other stakeholders involved that their fairness matters. Subject fairness is an instance of this type, where the goal is to be fair towards subjects of items being recommended or retrieved. For example, in news platforms, to avoid polarization in the society, we might want to have a fair representation of different points of views, or giving a fair coverage to different topics and avoiding certain popular topics monopolizing news feeds.
    
    % fairness over time
    To define, measure and incorporate fairness definitions in algorithms, we should take into account many different aspects of fairness as we mentioned above. Another important aspect to consider is measuring the dynamics of fairness. Recommendation engines, change over time, as they interact with their users, attract new users and lose other users in the process. Therefore achieving fairness in one iteration might not be enough and might overlook these temporal changes.
    As an example, in these systems, feedback loops might occur and this phenomena causes the system to pay more attention to popular/dominant subgroups of users \cite{hashimoto2018fairness} and therefore lose their under-represented sub-groups (either in consumers or providers). \cite{zhang2019group} analyzes the dynamics of fairness in sequential decision-making and tries to achieve a more balance performance which improves user retention. \cite{Chaney2018} studies how recommendations become more and more homogeneous in iterative environments which leads to inequity of exposure among items. 
    

% fairness for mitigation
    % fairness in data
    % fairness in ranking models
    % fairness through re-ranking
    % fairness through engineering

\subsection{achieving algorithmic fairness}
% achieving algorithmic fairness
% pre-processing methods
To achieve algorithmic fairness, interventions can be made at different steps of the processing pipeline. \cite{Friedler2019} provides a broad overview of these methods.

Pre-processing methods focus on compensating the existing biases in the dataset. \cite{chen2018why} suggest different data collection enhancements to compensate for the biases that occur due to data imbalance. In order to do so, if there is an under-represented group in the data, by collecting more data on that group or imputing the fundamental features of that group, we can see improvements in the performance metrics automatically. \cite{Feldman2015} modifies the numerical attributes in the data to equalize their marginal distributions conditioned on the sensitive attribute. In this way, these distributions will be independent of the sensitive attribute and therefore the outcome of the machine learning models which are trained on this data will be independent of the group membership. \cite{hajian2012methodology} suggest to modify the values of the attributes and labels in the data such that unfair association rules cannot be mined from the dataset. Some other approaches create intermediary (lower-dimensional) representations of the data points so as to hide the information about sensitive attributes, while keeping the utility
of the modified data for the required task \cite{zemel2013learning,lahoti2019ifair}. 
Despite all this work, there isn't much work that focuses on de-biasing the data in recommender systems. However, many of these work are applicable to the data that is appropriate for recommender systems. And are not sometimes enough to ensure the fairness of the outputs. Therefore the following methods are also considered.

%in-processing methods
In-processing approaches try to improve the fairness of results by modifying the algorithms and by integrating fairness notions in their loss functions. Therefore the problem will turn into a multi-objective optimization problem that seeks to simultaneously maximize utility and fairness.
These types of algorithmic interventions, sometimes take the form of regularizers to control certain structural properties of the model in the optimization functions. 
Although regularizers are usually used to control the complexity of the model and to prevent the model from overfitting, they can capture unfairness of the model as well. For example, \cite{zafar2017fairness} proposes to add fairness constraints on top of the accuracy constraints in the optimization objective of a classifier. Fairness regualrization has been used in other classification and regression problems such as \cite{kamishima2012fairness,berk2017convex} and recommendation problems. For example, \cite{kamishima2018recommendation, kamishima} proposes to add an independence term to the loss function that penalizes any correlations between the sensitive attribute and the predicted ratings. This term can also be added to achieve consumer-side fairness\cite{kamishima2017considerations}. They also propose multiple non-independence measures as well such as the difference in mean ratings between groups and the mutual information between the predicted ratings and the sensitive attributes. \cite{yao_huang_fatml-2017} also uses a regularization approach to minimize disparate rating predictions errors rather than recommendation errors. \cite{beutel2017data} adds a penalty term to their pairwise ranking loss function, to ensure that the difference between the ranking scores of the relevant and irrelevant items is uncorrelated with the relevant item's sensitive attribute. It is also possible to directly optimize a learning-to-rank such as \cite{diaz2020} that uses such method to achieve equal expected exposure.

Despite all the progress in this approach of increasing fairness, the goal of accuracy and fairness can contradict sometimes. Since traditionally a lot of these algorithms' objective functions are designed to achieve accuracy, adding fairness notions as an extra constraint might prevent the objective functions to converge. Therefore, post-processing methods provide alternatives where in-processing modification of algorithms is not fruitful.

%post-processing methods
Post-processing approaches, focus on modifying the outputs of the algorithms to satisfy a fairness criteria. In these methods, fairness constraints won't interfere with the goals of the objective function, rather they intervene after the output is produced. This approach can be applied both to classification and recommendation problems.
In \cite{fish2016confidence}, the proposed method tries to shift the boundaries of the already trained classifiers to achieve statistical parity with minimal accuracy loss. \cite{hardt2016equality} tries to balance the true positive rates of different groups by modifying the decision score thresholds of a trained classifier. \cite{kamiran2010discrimination} proposes a methodology that relabels the nodes of a decision tree classifier in order to ensure demographic parity.

Fairness can also be improved by re-ranking the output lists which were produced with the goal of achieving a high relevance. There are two main approaches of reranking: (a) those that treat the problem as a global optimization task and try to improve fairness with respect to the entire list of recommendations and (b) those methods that focus on the fairness of single lists one at a time.
An example of the first approach is \cite{surer2018multistakeholder} that proposes a constrained optimization-based method to enhance fairness (item exposure) for multiple provider groups, avoid unfairness towards under-represented groups and ensures a minimum degree of diversity for consumers. These methods are useful for occasion that recommendations are generated and cached in advance.
A more typical approach is to re-rank individual lists as they are generated. Such approaches use methods like MMR \cite{carbonell1998use} or xQuAD \cite{santos2010explicit} that were presented in the information retrieval literature. These methods propose a greedy list expansion approach, where the re-reranked list is generated by adding new items to this list where it satisfies a fairness or diversity criteria. This approach also provides the benefit of controlling the balance between the accuracy and the fairness goal. 
\cite{modani2017fairness} use a re-rank approach to enhance provider exposure while preserving relevance. \cite{Geyik2019} uses a greedy approach to produce rankings of job-candidates that have a fair distribution of their demographic attributes, simultaneously optimizing for fairness and relevance. \cite{zehlike2017fa} uses A-star algorithm for reranking to achieve fairness in a ranked list at depth K. The goal here is to re-arrange the ranked lists to meet a fair distribution of items from different protected groups while keeping the quality of ranked lists as high as possible. 




%then what we have done so far???
% in-processing
balanced neighbohood

% post-processing
far/pfar: defining a utility function for providers
ofair : contributions are having a multi-aspect fairness definition and constraint, opportunistic idea
dynamic fairness:

Their focus is on group fairness.